### 2017-05-15 lafaySimulatedScenePerception.tex
ml is done

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a software dedicated to sound edition, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points Likert scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
\Germanorfrenchabstract{}

@@

First, a subject can be asked to describe a given sound environnment following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, the amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, as those stimuli can be recordered and analysed.

@@

We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify the representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two above cited approaches. As the subject is asked to produces audio data (the signal of the simulated scene) allows the experimenter to study a modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.

@@

Recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}. For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very interesting outcomes as it allows the experimenter to conveniently study separately the influence of the sounds sources that compose a sound environment of interest. Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characteristics of the scene, that is the temporal distribution of the events.

@@

To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes. Results and outcomes of a series of 3 experiments that build on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:

@@

To the best of our knowledge, only Bruce~\al \cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources follows more social or semantic matters than their acoustical characteristics. A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. The three experiments,  1 .a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modification of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.

@@

L'expérience 1 .a a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.

@@

\emph{Simscene} is an online digital audio tool whose first version have been developped as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio} : \url{www.w3.org/TR/webaudio}}. The interface for selecting the soud sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

\emph{Simscene} is designed to be used as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sounds source within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.

@@

In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are available in urban areas, a typology of urban sounds is first agreed upon.

@@

The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures. These two types have specific simulation procedures.% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.

@@

We do so, because we assume that the morphological differences between those two classes have some important consequences on the perception of the scenes.

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes sperately events emerging from different sources of an sound environment \cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}. In this case, the source is a perceptual entity. Indeed, events that are not isolated are grouped into a stream.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events. She shows that those processes leads to two abstract cognitive category respectively termed ""~the event sequences ~" and "~ the amorphous sequences".

@@

The event sequences are composed of salient and identifiable events such as \emph{car start}, \emph{mal speech}. They arise from a descriptive analysis based on the identification of the sound sources. On contrary, the amorphous sequences are environments where distinct events can hardly be found, such as \emph{traffic hubhub}. They build upon an holistic analysis based on global acoustic features.

@@

Concerning sound textures, \ie~sounds that have stable caracteristics along time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrate that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.

Au regard de ces résultats, il apparaît que le processus de traitement de l'information sonore comprend une prise de décision quant à la nature des stimuli \cite{nelken2013ear,mcdermott2013summary}, laquelle va ensuite influer sur la manière d'extraire et d'analyser l'information.

At the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}. We thus consider in this study the soundscape as "skeleton of events on a bed of textures" as coined in \cite{nelken2013ear}.

@@

The sound classes are organized hierachically (\cf~Figure~\ref{fig:orgDb}) according to a structure close to the one of the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. In this taxonomy, the lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources. For class with high level of abstraction with sub classes, its collection of samples is the union of the collections of the sub classes.

@@

483 isolated sounds are collected and organized with the two above discussed typologies, 381 events and 102 textures. Among those samples, 332 have been recorded and 151 have been taken from two sound libraries: \emph{SoundIdeas}\footnote{\emph{SoundIdeas}: \url{www.sound-ideas.com}} and \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank}: \url{www.universal-soundbank.com}}.

@@

Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder. The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background. Conversely, it allowed us to avoid dominant sound sources while recording texture sounds by pointing the microphone to distant areas with no dominant sound sources.

@@

\begin{itemize}
\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each samples;
\item \emph{global fades} (seconds): the subject set the duration of the fade in /fade out at the beginning and the end of the track.
\end{itemize}

@@

Most of browsing tools are based on keyword indexing. For sensory experiment studying the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject. Indeed, the subject will select \textit{a priori} sound referenced belonging to \emph{park} to build a \emph{calm} soundscape.

@@

Therefore, the selection interface is text free and composed of spatially organized circles on a two dimensional plane.

@@

\begin{figure}[t]
        \myfloatalign
        \subfloat[]
        {\includegraphics[width=.8\columnwidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
        \subfloat[]
        {\includegraphics[width=\columnwidth]{gfx/ch_4/simscene}\label{fig:simscene}}
       \caption{Visual interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{Simscene} tool.}
\end{figure}

@@

As shown on Figure~\ref{fig:simscene}, the simulation interface displays a schematic of the scene under creation. Each track is represented as an horizontal band with a temporal axis. Each sample of this track is displayed by a rectangle whose height is proportional to the amplitude of the sample. For event's tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets. For tracks of texture, a unique rectangle is displayed as this kind of sounds do not allow spacing with silence. As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples. The subject can listen to the resulting sound scene.

@@

The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific of some of environments. Those sound markers seem to have a great impact on perception. This impact is studied in more detail using an added benefit of the simulation paradigm proposed in this paper, the ability to manipulate and modify the generated scenes.

@@

In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveform of the simulated scenes are regenerated without considering the classes identified as markers. For the i-scenes, the i-markers are removed. For the ni-scenes, the ni-markers are removed. A perceptive test evaluating the pleasantness is conducted with a protocol close to the one considered in experiment 1.b.

@@

Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are formulated:

@@

\begin{itemize}
\item \emph{Church bell}, \emph{bicycle bell}, \emph{animals} for the i/sm-scenes;
\item \emph{siren}, \emph{car horn} for the ni/sm-scenes.
\end{itemize}

@@

The subjects evaluate the 144 scenes. The evaluation is done on an 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant). Before grading a scene, subjects must listen to the first 20 seconds. After scoring, they are free to move on to the next scene.

For each subject, the scenes are presented in random order. The first 10 scenes allow the subject to calibrate his/her grades. Those calibration scenes are  5 i/km-scenes and 5 ni/km-scenes. These first 10 scenes are replayed at the end of the experiment, and only the notes given at the second occurrence are taken into account.

@@

The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness. To do so, we perform a variance analysis. We consider $\mathcal{A}_{sujet}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm). As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used in order to verify the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes. The two independent variables are considered as \emph{within-subject} factors. The factors being of only two levels each (type: i/ni; marker: km/rm), the sphericity hypothesis does not need to be checked. \emph{Post hoc} analysis are done by following the Tukey-Kramer procedure.

@@

Let us consider $\mathcal{A}_{sujet}$ for the km-scenes. Close inspection that the answers given by one subject strongly differs with the ones of others. This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7. This subject gave a grade above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the remaining of the subjects.

Furthermore, this subject used the whole scale (from -5 to 5) to grade both the i-scenes and the ni-scenes. Those behaviors strongly differing with the ones of the others subjects be they from this experiment or the previous ones, we decide to discard the subject 7 from the analysis.

@@

In this section, we study the grades given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes. The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ exhibit a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).

@@

So the i-markers do have a positive impact on the perception of an environment. The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantnes of a given place by the addition of sounds commonly considered as pleasant such as \emph{brid calls}. Those conclusions are in line with the positive approacj introduced by Schafer in \cite{schafer1977tuning}.

@@

These results tend to confirm that the perception of the qualities of a scene depends above all on the sound sources that compose it, the compostional characteristics taken into account during the perceptual process appearing to vary from one source to the other, from one type of environment to the other. This fact lead the authors to conclude that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.

@@

\begin{itemize}
\item in the case of non ideal scenes, one such focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
\item in the case of ideal scenes, one should first identity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
\end{itemize}

@@

First, the fact that the semantic information (in terms of which sound sources are present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the EA and ED concepts.

@@

Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics allowing to distinguish two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.

@@

In the opinion of the authors, the study described in this paper demonstrate the usefulness of considering simulation tools such as \emph{simScene} in order to scientifically question the perception of soundscapes. We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which it is exposed, and about the representations of the sound environments to which it would like to be exposed.

@@

It would also be interesting to use simulation in order to study further the effects caused by the voluntary modification of a characteristic of a scene, as during the suppression of the sound markers practiced in experiment 2.

@@

One interesting avenue in this vein would be to study the impact of adding positively appreciated sound markers to a non ideal scene in order to study the hypothesis that this type of addition would improve the perceived quality of the scene.

>>>>

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a software tool dedicated to sound edition, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points Likert scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
\Germanorfrenchabstract{}

@@

First, a subject can be asked to describe a given sound environnment following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, as those stimuli can be recordered and analysed, it allow the experimenter to gather a coarse grain physical description of the stimuli.

@@

We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify her representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two above cited approaches. As the subject is asked to produces audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.

@@

We believe that the availability of a fine grain description of the sound stimuli is of great insterest, beceause recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very interesting outcomes as it allows the experimenter to separately study  the influence of the sounds sources that compose a sound environment of interest. Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characteristics of the scene, that is the temporal distribution of the events.

@@

To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes. Results and outcomes of a series of three experiments that build on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:

@@

To the best of our knowledge, only Bruce~\al \cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. The three experiments,  1.a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modification of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}. Experiment 1.a is based on a pilot study described in \cite{lafay2014new}.

@@

%L'expérience 1.a a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.

@@

\emph{Simscene} is an online digital audio tool whose first version have been developped as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

\emph{Simscene} is designed as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sounds source within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.

@@

In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are available in urban areas, a typology of urban sounds is first defined.

@@

The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures. These two types have specific simulation procedures.

% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.

@@

We do so because we assume that the morphological differences between those two classes have some important consequences on the perception of the scenes.

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes seperately events emerging from different sources of a sound environment \cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}. In this case, the source is a perceptual entity as events that are not isolated are grouped into a stream.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events. She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".

@@

The event sequences are composed of salient and identifiable events such as \emph{car start}, \emph{male speech}. They arise from a descriptive analysis based on the identification of the sound sources. On contrary, the amorphous sequences are environments where distinct events can hardly be found, such as \emph{traffic hubhub}. They build upon an holistic analysis based on global acoustic features.

@@

Concerning sound textures, \ie~sounds that have stable caracteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.

%Au regard de ces résultats, il apparaît que le processus de traitement de l'information sonore comprend une prise de décision quant à la nature des stimuli \cite{nelken2013ear,mcdermott2013summary}, laquelle va ensuite influer sur la manière d'extraire et d'analyser l'information.

At the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}. We thus consider in this study the soundscape as a "skeleton of events on a bed of textures" as coined in \cite{nelken2013ear}.

@@

The sound classes are organized hierachically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources. For classes with high level of abstraction with sub classes, its collection of samples is the union of the collections of the sub classes.

@@

483 isolated sounds are collected and organized with the two typologies discussed above, 381 events and 102 textures. Among those samples, 332 have been recorded and 151 have been taken from two sound libraries: \emph{SoundIdeas}\footnote{\emph{SoundIdeas}: \url{www.sound-ideas.com}} and \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank}: \url{www.universal-soundbank.com}}.

@@

Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder. The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background. It also allowed us to avoid dominant sound sources while recording texture sounds by aiming the microphone to distant areas with no dominant sound sources.

@@

\begin{itemize}
\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each samples;
\item \emph{global fades} (seconds): the subject set the duration of the fade in / fade out at the beginning and the end of the track.
\end{itemize}

@@

Once the typology and the set sound are available, an important design question that remains is to find a suitable way of presenting the data to the user. Most of browsing tools are based on keyword indexing. However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject. Indeed, the subject may select \textit{a priori} sound referenced belonging to \emph{park} to build a \emph{calm} soundscape.

@@

Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a two dimensional plane.

@@

\begin{figure}[t]
        \myfloatalign
        \subfloat[]
        {\includegraphics[width=.8\columnwidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
        \subfloat[]
        {\includegraphics[width=\columnwidth]{gfx/ch_4/simscene}\label{fig:simscene}}
       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
\end{figure}

@@

As shown on Figure~\ref{fig:simscene}, the simulation interface displays a schematic of the scene under creation. Each track is represented as an horizontal band with a temporal axis. Each sample of this track is displayed by a rectangle whose height is proportional to the amplitude of the sample. For event's tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets. For tracks of texture, a unique rectangle is displayed as this kind of sounds do not allow spacing with silence. As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples. The subject can listen to the resulting sound scene at any time.

@@

The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific to some of environments. Those sound markers seem to have a great impact on perception. This impact is studied here in more detail using an added benefit of the simulation paradigm proposed in this paper, the ability to manipulate and modify the generated scenes.

@@

In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without considering the classes identified as markers. For the i-scenes, some of the i-markers are removed. For the ni-scenes, some of the ni-markers are removed. A perceptive test evaluating the pleasantness is then conducted with a protocol close to the one considered in experiment 1.b.

@@

Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are thus formulated:

@@

\begin{itemize}
\item \emph{Church bell}, \emph{bicycle bell}, and \emph{animals} for the i/rm-scenes;
\item \emph{siren}, \emph{car horn} for the ni/rm-scenes.
\end{itemize}

@@

The subjects evaluate the 144 scenes. The evaluation is done on a 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant). Before scoring a scene, subjects must listen to the first 20 seconds. After scoring, they are free to move on to the next scene.

For each subject, the scenes are presented in a random order. The first 10 scenes allow the subject to calibrate their scores. Those calibration scenes are  5 i/km-scenes and 5 ni/km-scenes. These first 10 scenes are replayed at the end of the experiment, and only the notes given at the second occurrence are taken into account.

@@

The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness. To do so, we perform a variance analysis. We consider $\mathcal{A}_{sujet}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm). As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to check the validity of the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes. The two independent variables are considered as \emph{within-subject} factors. The factors being of only two levels each (type: i/ni; marker: km/rm), the sphericity hypothesis does not need to be checked. \emph{Post hoc} analysis are done by following the Tukey-Kramer procedure.

@@

Let us consider $\mathcal{A}_{sujet}$ for the km-scenes. Close inspection that the answers given by one subject strongly differs with the ones of others. This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7. This subject also gave a score above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the remaining of the subjects.

Furthermore, this subject used the whole scale (from -5 to 5) to score both the i-scenes and the ni-scenes. Those behaviors strongly differing with the ones of the others subjects be they from this experiment or the previous ones, subject 7 is discarded from the analysis.

@@

In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes. The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ show a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).

@@

The i-markers do have a positive impact on the perception of an environment. The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantnes of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}. Those conclusions are in line with the positive approacj introduced by Schafer in \cite{schafer1977tuning}.

@@

These results tend to confirm that the perception of the qualities of a scene indeed depends above all on the sound sources that compose it. The compositional characteristics that are taken into account during the perceptual process appear to vary from one source to the other, from one type of environment to the other. This fact lead the authors to conclude that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.

@@

\begin{itemize}
\item in the case of non ideal scenes, one should focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
\item in the case of ideal scenes, one should first identity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
\end{itemize}

@@

First, the fact that the semantic information (in terms of which sound source is present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the EA and ED concepts.

@@

Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.

@@

The outcomes of the three experiments described in this paper demonstrate the usefulness of considering simulation tools such as \emph{simScene} in order to scientifically question the perception of soundscapes. We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which it is exposed, and about the representations of the sound environments to which it would like to be exposed.

@@

It would also be interesting to use the alteration capabilities of the proposed paradigm in order to study further the effects caused by the voluntary modification of a characteristic of a scene, as during the suppression of the sound markers practiced in experiment 2.

@@

One interesting avenue in this direction would be to study the impact of adding positively appreciated sound markers to a non ideal scene in order to study the hypothesis that this type of addition would improve the perceived quality of the scene.
